A Type System for Structured Aliasing in Parallel Programs

### 2012-10-29 doc/types-pldi2013/langdef.tex
relaxed coherence modes

\subsection{Relaxed Coherence}
\label{subsec:relaxed}

The analysis above does not consider the coherence modes of operations,
making the worst-case assumption that all operations use $excl$ mode.  However, if two or more
subexpressions contain calls to functions using $atomic$ or $simult$ coherence modes
on some regions, executions besides serial program order may 
result in valid interleavings even if constituent traces have locations in common.  We discuss
two cases of interest.  In both, the previous section's analysis 
is repeated with minor modifications, so for brevity we limit
discussion to a description of the differences rather than complete proofs.

The first important case occurs when all apparently-interfering operations have coherence modes
of either $atomic$ or $simult$.  In this case, the conditions required for
coherence and sequential equivalence are trivially satisfied for locations in $L_{excl}$.  The
requirements for locations in $L_{atomic}$ allow any permutation of the constituent traces,
so the runtime is free to serialize the execution of the subexpressions in any order, not just
program order.

To convert the required analysis to one that can be performed on regions, we modify
the non-interference operator for memory operations to consider only locations in $L_{excl}$ and show that
traces that are non-interfering under this modified check satisfy the requirement for atomic
execution.  We then modify the non-interference operator on privileges to allow 
$M(r_1) \cap M(r_2) \not= \emptyset$ if either $atomic(r_1)$ or $simult(r_1)$ is in $Q_1$ and
either $atomic(r_2)$ or $simult(r_2)$ is in $Q_2$. We then show that subexpressions with 
privileges that are {\em atomically non-interfering} (under a dynamic region mapping
$M$) always have memory traces compatible with atomically-reordered execution.  Finally,
in the special case where each subexpression has $atomic(r)$ or $simult(r)$ in $Q_i$ for every $r$
in $\Phi_i$, we can eliminate the dynamic check entirely.

The second case is a subset of the first, in which all apparently-interfering operations have
the $simult$ coherence mode.  In this case, all the coherence and sequential equivalence 
checks are trivially satisfied, and any interleaving of the constituent traces is valid.
We again modify the non-interference criteria for memory operations, traces, and privileges, this time
only allowing $simult$ coherence annotations. We show that subexpression with privileges
that are {\em simultaneously non-interfering} (under mapping $M$) can be executed
in parallel, which includes all the cases covered by Theorem~\ref{thm:parallelexec} as well
as those where the programmer has used the $simult$ coherence mode to explicitly allow parallel
execution.  Again, a special (but important) case exists where all privileges for each
subexpression are paired with $simult$ coherence mode declarations, allowing the safety of
parallel execution to be determined at compile time.

>>>>

\subsection{Atomic Coherence}
\label{subsec:coherenceatomic}

In cases where the Legion compiler and runtime cannot safely infer non-interference of
privileges (either because the programmer did not partition the regions enough or because the two
tasks really do access the same location), the fallback behavior of serialized execution can
be avoided through the use of relaxed coherence.

The first form of relaxed coherence is the {\em atomic} mode, which specifies that although two
tasks accessing the same region cannot run concurrently, they may execute in either order, allowing
a ``later'' task to run as soon as its ready, even if the ``earlier'' task is not ready.  This
relaxation only applies if all regions of overlap are designated as atomic coherence.

To show that this is safe, we define a relaxed version of non-interference for atomic coherence as
follows:
$$
\begin{array}{l@{}l}
op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \Leftrightarrow &op_1(l_1, c_1, v_1)\ \nonint^A\ op_2(l_2, c_2, v_2) \vee \\
& (c_1 = atomic \wedge c_2 = atomic)
\end{array}$$

We repeat the steps in Section~\ref{subsec:coherenceparallel} using the $\nonint^A$ operator and reach another
result used by the Legion runtime scheduler:

\begin{thm}
\label{thm:atomicexec}
\rm
Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
Let $M$ be a region mapping, $L$ a local value mapping,
$H$ a heap typing, and $S$ be an initial store satisfying
$\mapconsist{\Omega}$, $\localconsist{L}{\Gamma}$, and $\storeconsist{S}{H}$.
If $\Phi_i \nonint[M]^A \Phi_j$ for $1 \leq i < j \leq n$, then for any permuatation $(\oton{\pi})$ of
$(1, \ldots, n)$, $E_{\pi_1} \dplus \ldots \dplus E_{\pi_n}$ is a valid interleaving.
\end{thm}

\subsection{Simultaneous Coherence}
\label{subsec:coherencesimult}

The other way in which coherence can be relaxed is using the {\em simult} mode, which allows
multiple tasks to access the same region concurrently.  The simult coherence mode is appropriate in two
important cases:
\begin{enumerate}
\item When subtasks are accessing disjoint data, but the disjointness is hard to describe (e.g. walking
separate linked lists that have been allocated in the same region)
\item When the algorithm is tolerant of non-determinism (e.g. in a breadth-first search, setting the parent
pointer of a node with multiple equally-short paths to the root).
\end{enumerate}

When one or more regions have been marked as using the simult coherence mode, the non-interference operators from
the non-interference test is relaxed analagously to the atomic coherence case, yielding a $\nonint^S$ operator.
Because the rules for valid interleavings exclude locations that are only accessed in simult mode, it is straightforward
to extend Theorem~\ref{thm:parallelexec} to show that parallel execution is safe as long as $\Phi_i \nonint[M]^S \Phi_j$.

It is also possible to have both atomic and simult coherence modes in use at the same time (although for different regions),
in which case the non-interference test $\nonint^{AS}$ uses both the atomic and simult relaxations, and
Theorem~\ref{thm:atomicexec} is extended to allow arbitrary reordering (but not simultaneous execution) of subtasks when
$\Phi_i \nonint[M]^{AS} \Phi_j$.
